Finiteness and Cardinality in Homotopy Type Theory

### 2020-01-07 paper.tex
more!

\input{preamble}
\title{Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory}
\author{Donnacha Oisín Kidney}
\authorrunning{D. O. Kidney}
\institute{University College Cork \and
  \email{o.kidney@cs.ucc.ie}}
\begin{document}
\maketitle
\begin{abstract}
  We explore five notions of finiteness in Homotopy Type Theory \cite{hottbook}.
  We prove closure properties about all of these notions, culminating in a proof
  that decidable Kuratowski-finite sets form a topos.

  We extend the definitions to include infinite types, developing a similar
  classification of countable types.

  We use the definition of finiteness to formalise \emph{species}, in much
  the same way as in \cite{yorgeyCombinatorialSpeciesLabelled2014}.
  A clear duality with containers
  \cite{abbottContainersConstructingStrictly2005} falls out naturally from our
  definition.

@@

We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.

@@

For each predicate, we will also prove its closure over sums and products in
both dependent and non-dependent forms, if such a closure exists.
This will culminate in our main result for this section: the formal proof that
decidable Kuratowski finite sets form a topos.
More specifically, we show that cardinal finite sets form a topos, that
decidable Kuratowski finite sets are equivalent in strength to cardinal finite
sets, and carry the proof over.
This result relies on proofs on each of the other finiteness predicates.
\subsubsection{Species}
\todo[inline]{Redo this next paragraph}

\subsubsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the finiteness predicates are mirrored with countable
counterparts, and we will prove closure under the Kleene star and plus.

\subsubsection{Practical Uses of Finiteness}
Proofs of finiteness have well-known practical applications in
constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
In section~\ref{practical}, we build a library which exploits these uses in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
complex proofs over finite types.
We frame this in terms of the principle of omniscience for finite types.
Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
go further than the usual examples of this kind of proof automation: as well as
proving properties about functions, we can synthesise functions whole-cloth from
their desired properties.
Through the unified interface for finite and countable types, we can
reuse the automation machinery for \emph{partial} proof search over infinite
search spaces.
Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
prove isomorphisms between the curried and uncurried forms of \(n\)-ary
dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}

We will at this point take a moment to define some of the types we used to
define \(\mathcal{E}!\).
Lists, and membership thereof, are defined using \emph{containers}.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S \triangleright P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  \begin{equation}
    \begin{aligned}
      S : \text{Type}, P : S \rightarrow \text{Type} \\
      \mathbf{Container} = S \triangleright P
    \end{aligned}
  \end{equation}
  We ``interpret'' a container into a functor defined like so:
  \begin{equation}
    \llbracket S \triangleright P \rrbracket = \Pi {(X : \text{Type})} , \Sigma {(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{rm-definition}
Containers can be used to define a wide variety of functors (streams, trees,
etc.): lists are all that interest us now.
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Lists]
  The ``shape'' of lists is \(\mathbb{N}\), indicating the length of the list in
  question.
  \begin{equation}
    \mathbf{List} = \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}

Internally, in our formalisation, we actually use the standard
inductive definition of lists more often (it tends to work better in more
complex algorithms, and functions on it seems to satisfy the termination
checker more readily).
However, since both types are equivalent, univalence allows us to transport to
whichever representation is more convenient in a given situation.
For the higher-level proofs we present here, though, the container-based
definition greatly simplifies certain steps, which is why we have chosen it as
our representation.
\subsection{Split Surjections}
Another, equivalent way to define ``finiteness'' is via a (split) surjection
from a finite prefix of the natural numbers.
In this section, we will prove that equivalence, formally.

\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\end{proof}
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{rm-lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{rm-lemma}
  Any split enumerable type is a set.
\end{rm-lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{rm-lemma}
  \(\bot\), \(\top\), and \(\mathbf{Bool}\) are all split enumerable.
\end{rm-lemma}
\begin{proof}
  These non-recursive types have similar, simple proofs.
  For each we first provide the support list: they are \([]\), \([\text{tt}]\),
  and \([\text{false}, \text{true}]\) respectively.
  The cover proof should return an index which points at the given element: for
  \(\bot\) this function is present via the principle of explosion, for \(\top\)
  we always return a \(0\), and for \(\mathbf{Bool}\) we return a \(0\) for
  \(\text{false}\), and a \(1\) for \(\text{true}\).
  \begin{multicols}{3}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{unit}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{empty}
  \end{multicols}
\end{proof}
Next, we will look into how to combine proofs of split enumerability.
\begin{rm-theorem}
  Split-enumerability is closed under \(\sum\).
\end{rm-theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \Pi {(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-enum}
  ``do-notation'' is available to us as we're working in the list monad.
\end{proof}
\begin{rm-lemma}
  Split-enumerability is closed under disjoint union (non-dependent sum) and
  Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  Both of these closures can be derived from closure of the \(\sum\) type.
  The non-dependent product is equivalent to a special case of \(\sum\):
  \begin{equation}
    A \times B \simeq \Sigma {(x : A)} , B
  \end{equation}
  And disjoint union between two types \(A\) and \(B\) can be represented by the
  following type:
  \begin{equation}
    A + B = \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
  \end{equation}
  Then, since all of \(\mathbf{Bool}\), \(A\), and \(B\) are split enumerable,
  the type \(A + B\) is split enumerable.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction}; i.e. its membership proof
  exists, and all such proofs are equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
Where split enumerability was the enumeration form of a surjection from
\(\mathbf{Fin}\), we see here that manifest Bishop finiteness is the enumeration
form of an \emph{equivalence} with \(\mathbf{Fin}\).
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-lemma}
  Any manifest Bishop finite set is split enumerable.
\end{rm-lemma}
\begin{proof}
  The support set carries over simply, and the cover proof can be taken from the
  first component of the cover proof from the proof of manifest Bishop
  finiteness.
\end{proof}
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

  It suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
A defining feature of split enumerability and Bishop finiteness is decidability:
both predicates imply a decidable equality function on the underlying type.
To find a predicate for finiteness which doesn't imply this, and therefore
works with types other than just sets, we \emph{truncate} the membership proof.
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Propositional Truncation]a \label{prop-trunc}
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\).
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

@@

  We will use and consume values of the type \(\lVert A \rVert\) in three main
  ways.
  \begin{enumerate}
    \item \label{elim-prop-prop} We can first eliminate from \(\lVert A \rVert\)
      into any type which is a proposition.
      In other words, given a function \(f : A \rightarrow B\), and a proof that
      \(B\) is a proposition, we can construct a function \(\lVert A \rVert
      \rightarrow B\).
    \item \label{elim-prop-monad} As a consequence of this first point, we can
      always eliminate into another propositionally truncated type.
      As a result, \(\lVert \cdot \rVert\) forms a Monad: for our purposes, this
      simply means tat we can work ``under'' a propositional truncation in an
      ergonomic way.
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\).
      \begin{equation}
        \Pi {(x , y : A)} , f(x) \equiv f(y)
      \end{equation}
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{rm-definition}

By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\begin{proof}
  As with the other surjection proof (lemma~\ref{split-enum-surj}), this is
  simply a reassociation.
\end{proof}
\subsection{Relation to Split Enumerability}
\begin{rm-lemma}
  Any split enumerable type is also manifestly enumerable.
\end{rm-lemma}
\begin{proof}
  The proof carries over via truncation of the cover proof.
\end{proof}
\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

@@

  Kuratowski finite sets are closed under \(\sum\).

@@

  This follows from the \(\sum\) closure proof on manifestly enumerable types.

@@

\section{Species and Containers} \label{species-and-containers}
\todo[inline]{Figure out this section}

>>>>

\input{preamble}
\title{Finiteness and Cardinality in Homotopy Type Theory}
\author{Donnacha Oisín Kidney}
\authorrunning{D. O. Kidney}
\institute{University College Cork \and
  \email{o.kidney@cs.ucc.ie}}
\begin{document}
\maketitle
\begin{abstract}
  We explore five notions of finiteness in Homotopy Type Theory \cite{hottbook}.
  We will classify these notions, and prove their closure properties,
  culminating in a proof that decidable Kuratowski-finite sets
  \cite{kuratowskiNotionEnsembleFini1920} form a topos.

  We then develop a parallel classification for the countably infinite types, as
  well as a proof of the countability of \(A^\star\) for a countable type \(A\).

@@

We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.
Our main proof in this regard is that we can derive manifest Bishop finiteness
from Kuratowski finiteness plus a total order.

@@

For each predicate, we will also prove its closure over sums and products in
both dependent and non-dependent forms, if such a closure exists.
This will culminate in our main result for this section: the formal proof that
decidable Kuratowski finite sets form a topos.
More specifically, we show that cardinal finite sets form a topos, that
decidable Kuratowski finite sets are equivalent in strength to cardinal finite
sets, and carry the proof over.
This result relies on proofs on each of the other finiteness predicates.

\subsubsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the classification of finiteness predicates mirrors that of
their countable counterparts, and we will prove closure under the Kleene star.

\subsubsection{Practical Uses}
Finally, in section~\ref{practical} we will demonstrate some practical uses of
the finiteness proofs in Cubical Agda.
We will show how to use well-known techniques
\cite{firsovDependentlyTypedProgramming2015, devrieseBrightSideType2011} to
automate proofs of functions with finite (and infinite) domains.
We further show how to automate the synthesis of functions from desired
properties.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.

Before giving the definition of the predicate, we will first define lists (and
membership thereof): we have chosen a \emph{container}-based definition for this
work.
\begin{rm-definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a pair
  \(S \triangleright P\) where \(S\) is a type, the elements of which are called
  the \emph{shapes} of the container, and \(P\) is a type family on \(S\), where
  the elements of \(P(s)\) are called the \emph{positions} of a container.
  We ``interpret'' a container into a functor defined like so:
  \begin{equation}
    \llbracket S \triangleright P \rrbracket(A) = \Sigma {(s : S)} , \left( P(s) \rightarrow A \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
  Where \(\text{fiber}\) is from \cite[definition 4.2.4]{hottbook}.
\end{rm-definition}
\begin{rm-definition}[Lists]
  \begin{equation}
    \mathbf{List} = \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[\mathbf{Fin}]
  \(\mathbf{Fin}\)(n) is the type of natural numbers smaller than \(n\).
  It is defined inductively, as follows:
  \begin{equation}
    \begin{aligned}
      \mathbf{Fin}(0) &= \bot \\
      \mathbf{Fin}(n + 1) &= \top + \mathbf{Fin}(n)
    \end{aligned}
  \end{equation}
\end{rm-definition}

Internally, in our formalisation, we actually use the standard
inductive definition of lists more often (it tends to work better in more
complex algorithms, and functions on it seems to satisfy the termination
checker more readily).
However, since both types are equivalent, univalence allows us to transport to
whichever representation is more convenient in a given situation.
For the higher-level proofs we present here, though, the container-based
definition greatly simplifies certain steps, which is why we have chosen it as
our representation.

Finally, we can define formally split enumerability.
\begin{rm-definition}[Split Enumerable Set]
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\subsection{Split Surjections}
\begin{rm-definition}[Surjections]
  \begin{multicols}{2}
      A \twoheadrightarrow B = \Sigma (f : A \rightarrow B) , \text{surjective}(f)
    \end{equation} \break
      A \twoheadrightarrow! \; B = \Sigma (f : A \rightarrow B) , \text{split surjective}(f)
  \end{multicols}
  Where the definitions of surjective are taken from \cite[definition
  4.6.1]{hottbook}.
An equivalent way to define ``finiteness'' is via a split surjection from a
finite prefix of the natural numbers.
\begin{rm-theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
    \mathcal{E}!(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
  To be clear: in Agda, the proof could simply be \(\AgdaFunction{reassoc}\); we
  have written out the extra lines for clarity alone.
\subsection{Decidable Equality}
\begin{rm-lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, we know that
this implies split enumerable types must all be sets.
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
In order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
  \(\mathbf{Bool}\), \(\top\), and \(\bot\) are all split enumerable.
  These non-recursive types have similar, simple proofs.
  \begin{multicols}{3}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{bool}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{unit}
    \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{empty}
  \end{multicols}
  Split-enumerability is closed under \(\Sigma\).
    \frac{
      \mathcal{E}!(A) \; \; \; \Pi (x : A) , \mathcal{E}!(U(x))
    }{
      \mathcal{E}!(\Sigma (x : A) , U(x))
    }
  To obtain the support list, we concatenate the support lists of all the proofs
  of split-finiteness for \(U\) over the support list of \(E_A\).
  In Agda:
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{sigma-enum}
  ``do-notation'' is available to us as we're working in the list monad.
Closure under disjoint union and Cartesian product both follow from \(\Sigma\),
as these types can be defined in terms of \(\Sigma\).
\begin{align}
  A \times B &= \Sigma {(x : A)} , B \\
  A + B &= \Sigma {(x : \mathbf{Bool})} , \text{if} \; x \; \text{then} \; A \; \text{else} \; B
\end{align}
\section{Manifest Bishop Finiteness} \label{manifest-bishop}
Where split enumerability was the enumeration form of a split surjection from
\(\mathbf{Fin}\), manifest Bishop finiteness is the enumeration form of an
\emph{equivalence} with \(\mathbf{Fin}\).

\begin{rm-definition}[Manifest Bishop Finiteness]
  \begin{equation}
    \mathcal{B}(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \inunique xs
  \end{equation}
\end{rm-definition}
The only difference between this predicate and split enumerability is the list
membership term: we use \(\inunique\) here, where \(x \inunique \mathit{xs}\) is
to be read as ``\(x\) occurs exactly once in \(\mathit{xs}\)''.
\begin{rm-definition}[Unique Membership]
  We say an item \(x\) is ``uniquely in'' some container \(\mathit{xs}\) if its
  membership in that list is a \emph{contraction} \cite[definition
  3.11.1]{hottbook}; i.e. its membership proof exists, and all such proofs are
  equal.
  \begin{equation}
    x \inunique \mathit{xs} = \text{isContr}(x \in \mathit{xs})
  \end{equation}
\end{rm-definition}
By the definition of a contraction, we can always recover the underlying
membership proof, meaning that we can always derive split enumerability from
manifest Bishop finiteness.

A nice consequence of prohibiting duplicates is that now the length of the
support list is the same as the cardinality of the set.
\subsection{Equivalence}
\begin{rm-lemma} \label{bishop-equiv}
  A proof of manifest Bishop finiteness is equivalent to an equivalence with a
  finite prefix of the natural numbers.
  \begin{equation}
    \mathcal{B}(A) \simeq \Sigma {(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \simeq A \right)
  \end{equation}
\end{rm-lemma}
\begin{proof}
  There are many equivalent definitions of equivalence in HoTT.
  Here we take the version preferred in the Cubical Agda library: contractible
  maps \cite[definition 4.4.1]{hottbook}.
  Because of the parallels between contractible maps and split surjections,
  the proof proceeds much the same as \ref{split-enum-surj}.
  In other words, the definition of Bishop finiteness is itself a reassociation
  of a contractible map.
\end{proof}
\subsection{Relationship to Split Enumerability}
We now show that manifest Bishop finiteness has equal strength to split
enumerability.
\begin{rm-theorem} \label{split-enum-to-manifest-bishop}
  Any split enumerable set is manifest Bishop finite.
\end{rm-theorem}
\begin{proof}
  Let \(E\) be a proof of split enumerability for some set \(A\).
  From proposition~\ref{split-enum-discrete} we can derive decidable equality on
  \(A\), and using this we can define a function \(\text{uniques}\) which filters
  out duplicates from lists of \(A\)s.
  \begin{equation}
    \text{uniques} : \textbf{List}(A) \rightarrow \textbf{List}(A)
  \end{equation}
  This gives us our support list.

  It suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.
\end{proof}
Note that while manifest Bishop finiteness and split enumerability are equivalent
in ``strength'', the two types are not equivalent.
For instance, for a type \(A\) with \(n\) inhabitants, there are infinitely many
inhabitants of \(\mathcal{E}!(A)\), while there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
Both split enumerability and manifest Bishop finiteness are restricted to sets:
because they both imply decidable equality, no non-set types can satisfy them as
predicates.
To find a more general predicate for finiteness, we \emph{truncate} the
membership proof.
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Propositional Truncation]a \label{prop-trunc}
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\) \cite[3.7]{hottbook}.
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

@@

  We will use and consume values of the type \(\lVert A \rVert\) in three main
  ways.
  \begin{enumerate}
    \item \label{elim-prop-prop} We can first eliminate from \(\lVert A \rVert\)
      into any type which is a proposition.
      In other words, given a function \(f : A \rightarrow B\), and a proof that
      \(B\) is a proposition, we can construct a function \(\lVert A \rVert
      \rightarrow B\).
    \item \label{elim-prop-monad} As a consequence of this first point, we can
      always eliminate into another propositionally truncated type.
      As a result, \(\lVert \cdot \rVert\) forms a Monad: for our purposes, this
      simply means that we can work ``under'' a propositional truncation in an
      ergonomic way.
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\).
      \begin{equation}
        \Pi {(x , y : A)} , f(x) \equiv f(y)
      \end{equation}
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{rm-definition}

We will use the circle as an example non-set type which is nonetheless
manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\begin{proof}
  As with the other surjection proof (lemma~\ref{split-enum-surj}), this is
  simply a reassociation.
\end{proof}
\subsection{Relation to Split Enumerability}
\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

@@

  Kuratowski finite sets are closed under \(\Sigma\).

@@

  This follows from the \(\Sigma\) closure proof on manifestly enumerable types.

@@
